\addchap{\lsAcknowledgementTitle} 

I thank Hubert Haider, Matthias Hüning, and Shravan Vasishth for discussion and Martin Haspelmath and Tibor Kiss for detailed comments
on an earlier version of this book. I thank Antonio Machicao y Priemer for comments on sections
about alternative approaches. I thank Shravan Vasishth for discussion of psycholinguistic issues.

I want to thank all students of my course on the structure of Germanic languages. This book
benefited enormously from their questions and the discussion during the lectures. 
Carolin Ulmer %05.09. WALS falsch zitiert.
and
Bastian Schoenfeld % 14.06.2020 zur Behauptung, dass die Argumente im Englischen alle vollständig
                   % sein müssen, was für VPen nicht stimmt.
deserve special mention for important comments. I thank 
Sarah Böke,
Alexandra Fosså,
Anne Kilgus,
Nils P. Kujath
and 
Maya Tabea Wolff  
for pointing out typos. I thank Hannah Schleupner for help with Figure~\ref{fig-history-relations-germanic}.

I thank the Department of English Linguistics at the University of Frankfurt/""Main for inviting me to
talk about the approach to Germanic syntax that is the basis for this book.

I want to thank all proofreaders (\makeatletter\@proofreader\makeatother), who did much more than
usual copy editing and provided feedback on 
content as well.  Spezial thanks go to Brett Reynolds, and Anne Zaenen.

This book was partly funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – SFB 1412, 416591334, Project A04.

I thank the Twitter community for discussing gender issues and names in linguistics examples with
me. Special thanks go to Lauren Ackerman, Kirby Conrod, and Brianna Wilson for extensive discussion.





%      <!-- Local IspellDict: en_US-w_accents -->
